My First R Markdown File
========================

<!-- Comments 
Setting global options
an example -->
<!-- ```{r setoptions, echo=False} -->
<!-- opts_chunk$set(echo=FALSE, results ="hide") -->
<!-- ``` -->

This is my first R markdown dile.

Here, I'm going to load some data.

```{r}
library(datasets)
data(airquality)
summary(airquality)
```

Make a pairs plot of the data.

```{r}
pairs(airquality)
```

Regression model of ozone on wind, solar radiation and temperature

```{r}
library(stats)
fit <- lm(Ozone ~ Wind + Solar.R + Temp, data = airquality)
summary(fit)
```

Unordered list:

 • First element
 
 * Second element
 

This is some text (i.e. a "text chunk")

Here is a code chunk
```{r}
set.seed(1)
x <- rnorm(100)
mean(x)
```

Another example

# My First kniter Document
Mariana S.

## Introduction

This is some text (i.e. a "text chunk").
Here is a code chunk (it wont be echoed - wont appear in the document generated)
```{r simulation, echo=FALSE}
set.seed(1)
x <- rnorm(100)
mean(x)
```

This is some text (i.e. a "text chunk").
Here is a code chunk but it does not print anything
```{r simulation2, echo=FALSE, results="hide"}
set.seed(1)
x <- rnorm(100)
mean(x)
```

## Inline Text Computations
### Intro
```{r computetime,echo=FALSE}
Sys.setenv(TZ="Europe/London")
Sys.getenv("TZ")
time <- format(Sys.time(), "%a %b %d %Y %X ")
rand <- rnorm(1)
```

The current time is `r time`.
My favourite random number is `r rand`.


## Incorporating GRaphics
### Intro
Simulate some data.
```{r simulatedata, echo=TRUE}
x <- rnorm(100)
y <- x + rnorm(100, sd = 0.5)
```

Here is a scatterplot of the data.
```{r scatterplot, fig.height=4}
par(mar = c(5,4,1,1), las = 1)
plot(x, y, main = "My Simulated Data")
```

## Making Tables with xtable
### Intro

```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone ~ Wind + Temp + Solar.R, data = airquality)
```

Here is a table of regression coefficients.
```{r showtable, results="asis"}
library(xtable)
xt <- xtable(summary(fit))
print(xt, type = "html")
```






